---
title: "ESM 244: Lab 1"
author: "Julianna Renzi"
date: "1/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here) # really nice paths to read in files
library(sf) # spatial data wrangling
library(tmap) # interactive map

```

### Read in San Francisco trees data

```{r}

sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv")) # gives relative file path from project root

```

### Basic wrangling reminders

Refresh some skills for data wrangling and summary stats using functions in the 'dplyr' package. Note can use Help -> Cheatsheets for more foramtting options.

Find the top 5 highest observations of trees by legal_status. Then do some wrangling and make a graph.

```{r}
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% # drop NAs where legal_status is NA
  rename(tree_count = n) %>% # new name on left
  relocate(tree_count) %>% # to re-order (e.g. by number/characters, or in this case move one to the front)
  slice_max(tree_count, n = 5) # allows you to get max values of a variable and keep those above a threshold (i.e. top 5 in this case)

# count trees observed by legal status
# press enter after each pipe operator
# count() counts observations by groups (recognizes groups, finds counts, and puts it in a table--group_by, summarize n() all in one!) 

sf_trees %>% 
  count(legal_status, species) # can do multiple for example


```

Make a graph of those top 5 observations by legal status

```{r}

ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + 
  geom_col() + # column graph (default is alphabetical so sort)
  # convert to factor and reorder based on tree count
  labs(x = "Legal Status", y = "Tree Count") + 
  coord_flip() + # flip axes for better text
  theme_minimal()

```

### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% # keep all observations where the string is detected anywhere in the cell (look in sf_trees, keep in spp variables the string "Blackwood Acacia" is detected)
  select(legal_status, date, latitude, longitude) # clean it up and remove columns 

# map it!
ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) +
  geom_point()

```

### tidyr::separate() and unite() functions

Useful for combining or separating columns

```{r}
# split by colon
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = "::") 

```

Example: tidyr::unite()

```{r}
# combine columns and separator
sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep = "_cool_") # unite a series of columns with a common separator and call it id_status
```

## Make actual maps

'st_as_sf()' to convert lat & long to spatial coordinates

```{r}
# convert to spatial points
blackwood_acacia_sp <- blackwood_acacia %>% 
  drop_na(longitude, latitude) %>% # remove any na's for lat and long
  st_as_sf(coords = c("longitude", "latitude")) # what are the variables storing long/lat -- order MATTERS (put lon then lat variables)
  # now have a geometry column--can wrangle columns but geometry will stay with rows until we tell it not to ('sticky')

# assign coordinate system
st_crs(blackwood_acacia_sp) = 4326 # European Petroleum Survey group code for coordinate references

# now can plot spatial data
ggplot(data = blackwood_acacia_sp) + 
  geom_sf(color = "darkgreen") + 
  theme_minimal() # but hard to interpret without san francisco shape file

```

Read in SF roads shapefile:

```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

# check
st_transform(sf_map) # has bounding box (bbox), etc. -- want them to be in the same Coordinate Reference System (CRS)--there is an existing crs here!

# transform it so it's on the same crs
st_transform(sf_map, 4326)

ggplot(data = sf_map) +
  geom_sf(size = .1) # San Francisco roads

```

Combine blackwood acacia observations and SF roads

```{r}
ggplot() +
  geom_sf(data = sf_map, size = 0.1, color = "darkgray") +
  theme_void() + 
 geom_sf(data = blackwood_acacia_sp, color = "red", size = 0.5) 
```

Now create an interactive map

```{r}
tmap_mode("view") # default is plot (static)--want it to be interactive!

# create the tm object
tm_shape(blackwood_acacia_sp) + # needs a crs specified
  tm_dots() # shape
# WOAH!!!
  
```

